Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-06 paper/main.tex
Updates

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations} 
\todo[inline]{Needs improvement}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.

@@

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous variables, Pearson's correlation
coefficient is typically used. In the case when $ \bm{Z} = \emptyset $, the
correlation coefficient is defined as:

@@

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient
can be used. One of the ways to estimate it is to use two regression
models $ E_X: X \sim \bm{Z} $, $ E_Y: Y \sim \bm{Z} $, and compute the Pearson's 
correlation coefficient between the residuals of these regression models:

\begin{equation}
	\begin{split}
		R_X &= X - E_X(\bm{Z})  \;\;\;\;\;\;\;\;  R_Y = Y - E_Y(\bm{Z}) \\
		\rho_{X, Y; \bm{Z}} &= \frac{\mathrm{cov}(R_X, R_Y)}{\sigma_{R_X} \sigma_{R_Y}}
	\end{split}
\end{equation}

@@

In the case of all discrete variables, Cram\'er's V can used as a measure of
association. When $ \bm{Z} = \emptyset $, Cram\'er's V can be computed from the 
chi-squared test statistic as:

\begin{equation}
	\mathbb{V_{X, Y}} = \sqrt{\frac{\chi^2 / n}{\mathrm{min}(k-1, r-1)}}
\end{equation}

where $ \chi^2 $ is the chi-squared test statistic for $ X $ and $ Y $, $ n $ 
is the sample size and $ k $ and $ r $ are the number of categories of $ X $ and 
$ Y $.

However, when $ \bm{Z} \neq \emptyset $, we first split the dataset by unique
values of $ \bm{Z} $, $ D = \{D_{\bm{Z} = \bm{Z_1}}, D_{\bm{Z} = \bm{Z_2}},
\cdots, \} $  and then combine them as:

\begin{equation}
	\mathbb{V_{X, Y; Z}} = \sum_{D_{\bm{Z} = \bm{Z_i}}} \mathbb{V}_{X_{D_i}, Y_{D_i}}
\end{equation}

\paragraph{$ X $ is ordinal and $ Y $ is continuous or ordinal}
When one of the variables is ordinal and the other is continuous polyserial
correlation can be used. When both the variables are ordinal polychoric
correlation can be used. Both polychoric and polyserial correlation make the
assumption that the ordinal variable has an underlying normal distributed
continuous variable, and the ordinal variable has been generated by
thresholding this latent variable. The method then tries to maximize the
likelihood of data given polychoric or polyserial correlation coefficient. This 
method gives us the covariance matrix, $\Sigma$ on the variables $ X $, $ Y $, and $ \bm{Z} $.

\begin{equation}
	\rho_{X, Y} = \Sigma_{XY}
\end{equation}

@@

In the previous section, we discussed how different types of variables require
different measures of conditional association. However, there is no unified
measure that handles mixed data. Here, we introduce a measure of conditional
association for mixed data by extending the concept of partial correlation
coefficient (commonly used for continuous variables) to mixed data. Similar to
partial correlation coefficient, our method integrates a mixed data
residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, our goal is to estimate the conditional association $ \rho(X, Y; \bm{Z}) $. 
In the first step, we compute the residuals $ R_X $ and $ R_Y $ for variables
$ X $ and $ Y $ respectively. Depending on the type of variable the residual
is computed as follows:

@@

The second step is to quantify the association between these residual matrices.
For this purpose we use canonical correlations that have been widely used to
measure the association between sets of random variables.

\begin{definition}
	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\rho(\bm{U}, \bm{V}) $ is defined as:

	\begin{equation}
		\rho(\bm{U}, \bm{V}) = \max_{a, b} Corr(a^T \bm{U}, b^T \bm{V})
	\end{equation}

@@

Several measures of association have been derived from canonical correlations, such as:
\begin{itemize}
	\item Wilks' Lambda: $\Lambda = \prod_{i}^{\min(p, q)} (1 - \phi_i^2) $
	\item Roy's Largest Root: $ \theta = \max_i(\phi_i^2) $
	\item Pillai's Trace: $ \tau = \sum_{i=1}^{\min(p, q)} \phi_i^2 $
\end{itemize}

We use a slightly modified version of Pillai's Trace for our purpose for two
main reasons: 1) It uses all the canonical correlation values, capturing the
full extent of association, 2) Its interpretation is similar to Pearson's
correlation coefficient, i.e., $ 0 $ signifies no association and $ 1 $
signifies perfect linear relationship. This leads to our proposed measure of
conditional association

\begin{equation}
	\rho(X, Y; \bm{Z}) = \frac{1}{\min(\rvert R_X \rvert, \rvert R_Y \rvert)}
	\sum_{i=1}^{\min(\rvert R_X \rvert, \rvert R_Y \rvert)} (\phi_i(R_X, R_Y))^2
\end{equation}

>>>>

\section{Background and Related Work}
\label{sec:background}
\subsection{Notations} 
\todo[inline]{Needs improvement}
We denote random variables with uppercase letters $ X $, and a set of random
variables as $ \bm{X} = \{X_1, \cdots, X_k\} $ with $ \rvert \bm{X} \rvert = k
$. A sample from the random variable $ X $ is denoted as $ x $ and from a set
of random variables $ \bm{X} $ is denoted as $ \bm{x} $. In this paper, we
consider random variables in the mixed data setting where each of the random
variables can be either of continuous, categorical, or ordinal unless
specified. We write the expectation of a variable $ X $ as $ \mathbb{E}[X] $,
conditional expectations as $ \mathbb{E}[X | \bm{Z}] $. We denote a graph $ G =
(V, E) $ which nodes $ V$ and edges $ E $. In this paper we focus on all
observed variables, and linear measure of association. 
where $ \mathrm{cov}(X, Y) $ is the covariance between $ X $ and $ Y $, $ \mathrm{corr}(X, Y) $ is the correlation between $ X $ and $ Y $, and $
\sigma_X $ and $ \sigma_Y $ are the standard deviations of $ X $ and $ Y $
respectively.  Covariance matrix using $ \Sigma $ and the entry corresponding to
covariance between $ X $ and $ Y $ in the matrix using $ \Sigma_{XY} $.

@@

\subsubsection{Both $ X $ and $ Y $ are continuous}
When both $ X $ and $ Y $ are continuous, Pearson's correlation coefficient is
typically used. When $ \bm{Z} = \emptyset $, the correlation
coefficient is defined as:

@@

When $ \bm{Z} \neq \emptyset $, partial correlation coefficient can be used.
This is estimated by fitting two regression models $ E_X: X \sim \bm{Z} $ and $
E_Y: Y \sim \bm{Z} $, calculating the residuals $ R_X = X - E_X(\bm{Z}) $ and $
R_Y = Y - E_Y(\bm{Y}) $, and then computing the correlation between the
residuals:

\begin{equation}
	\rho_{X, Y; \bm{Z}} = \rho_{R_X, R_Y}
\end{equation}

@@

When all variables are discrete, Cram\'er's V can used as a measure of
association. When $ \bm{Z} = \emptyset $, Cram\'er's V is derived from the 
chi-squared test statistic:

\begin{equation}
	\mathbf{V}_{X, Y} = \sqrt{\frac{\chi^2 / n}{\mathrm{min}(k-1, r-1)}}
\end{equation}

where $ \chi^2 $ is the chi-squared test statistic for $ X $ and $ Y $, $ n $
is the sample size, and $ k $ and $ r $ are the number of categories in $ X $
and $ Y $ respectively. When $ \bm{Z} \neq \emptyset $, we start by splitting
the dataset by each unique combination of $ \bm{Z} $ into subsets,  $ \bm{Z} $,
$ D = \{ D_{\bm{Z} = \bm{Z}_1}, D_{\bm{Z} = \bm{Z}_2}, \cdots, D_{\bm{Z} =
\bm{Z}_k} \} $. Then we compute the Cram\'er's V for each of these smaller datasets
and combine them as follows:

\begin{equation}
	\mathbf{V}_{X, Y; \bm{Z}} = \sum_{i=1}^{k} \left[ \mathbf{V}_{X, Y} \right]_{D_{\bm{Z} = \bm{Z}_i}} 
\end{equation}

\paragraph{$ X $ is ordinal, and $ Y $ and $ Z $  are continuous or ordinal}

Polyserial (when one is ordinal and the other is continuous) and Polychoric
(when both are ordinal) correlation have been used to estimate the covariance
matrix between them \citep{Poon1987}. Both methods make the assumption that the
observed ordinal variable is a result of thresholding a latent normally
distributed continuous variable. Under this assumption, the methods then try to
estimate the covariance matrix while maximizing the likelihood of the dataset.
Using the estimated covariance matrix, $ \Sigma $ we can the compute Pearson's
correlation coefficient. When $ \bm{Z} = \emptyset $,

\begin{equation}
	\rho_{X, Y} = \frac{\Sigma_{XY}}{\sqrt{\Sigma_{XX} \Sigma_{YY}}}
\end{equation}

@@

In the previous section, we discussed how different types of variables require
different measures of conditional association. However, there is no unified
measure that handles mixed data. In this section, we introduce a measure of conditional
association for mixed data by extending the concept of partial correlation
coefficient (commonly used for continuous variables) to mixed data. Similar to
partial correlation coefficient, our method integrates a mixed data
residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, our goal is to estimate the conditional association $ \phi_{X, Y; \bm{Z}} $. 
In the first step, we compute the residuals $ R_X $ and $ R_Y $ for variables
$ X $ and $ Y $ respectively. Depending on the type of variable the residual
is computed as follows:

@@

The second step is to quantify the association between these residual matrices.
For this purpose we use canonical correlations \citep{Hotelling1936} that have been widely used to
measure the association between sets of random variables.

\begin{definition}
	Given two sets of random variables $ \bm{U} = (U_1, U_2, \cdots, U_p) $
	and $ \bm{V} = (V_1, V_2, \cdots, V_q) $, canonical correlation between
	them, $\upsilon(\bm{U}, \bm{V}) $ is defined as:

	\begin{equation}
		% \nu_{\bm{U}, \bm{V}}= \max_{a, b} \mathrm{corr}(a^T \bm{U}, b^T \bm{V})
		\rho_{\bm{U}, \bm{V}} = \max_{a, b} \frac{a^T \Sigma_{\bm{UV}} b}{\sqrt{a^T \Sigma_{\bm{UU}} a \cdot b^T \Sigma_{\bm{VV}} b}}
	\end{equation}

@@

Several measures of association have been derived from canonical correlations, such as:
\begin{itemize}
	\item Wilks' Lambda: $\Lambda = \prod_{i}^{\min(p, q)} (1 - \nu_i^2) $
	\item Roy's Largest Root: $ \theta = \max_i(\nu_i^2) $
	\item Pillai's Trace: $ \tau = \sum_{i=1}^{\min(p, q)} \nu_i^2 $
\end{itemize}

We use a slightly modified version of Pillai's Trace for our purpose for two
main reasons: 1) It uses all the canonical correlation values, capturing the
full extent of association, 2) Its interpretation is similar to Pearson's
correlation coefficient, i.e., $ 0 $ signifies no association and $ 1 $
signifies perfect linear relationship. This leads to our proposed measure of
conditional association: We divide to make the measure normalized when comparing
between pair of variables with different number of categories.

\begin{equation}
	\rho(X, Y; \bm{Z}) = \frac{1}{\min(\rvert R_X \rvert, \rvert R_Y \rvert)}
	\sum_{i=1}^{\min(\rvert R_X \rvert, \rvert R_Y \rvert)} (\nu_i(R_X, R_Y))^2
\end{equation}
